---
title: ""
author: ""
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: ["book.bib",  "packages.bib"]
biblio-style: "apalike"
csl: apa.csl
header-includes:
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr'
), 'packages.bib')
```


```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1. Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported ('yes') when the response is signficant in univariate models and included in all top multivariate models and as partially supported ('yes') when the direction of response consistently matched the prediction but  the effect was not significant in all models.") %>%
  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("1a. Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("1b. Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("1c. Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("2. Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1.2cm") %>%
  column_spec(6, width = "2.4cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kableExtra::landscape()
```
\clearpage

 ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
  library(knitr)
  library(kableExtra)
  table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
  kable(table2, booktabs = TRUE, caption = "**Table 2. Summary of variables**") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
  column_spec(4, width = "7cm") %>%
  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  kable_styling()
```
\clearpage

```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table3 <- read.csv("tables_figures/table3_species_table.csv")

kable(table3, booktabs=TRUE, caption = "**Table 3. Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
*Semi-ring porosity is intermediate between ring and diffuse. We group it with diffuse-porous species for more even division of species between categories.

\clearpage
```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
univar_table <- read.csv("tables_figures/table4_univariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(univar_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
  kable_styling()
```

\clearpage
```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/table5_multivariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(multivar_table, booktabs = TRUE, caption = "**Table 5. Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 2," ", " ")) %>%
  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 5, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 6, 9, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 10, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(5, bold = T) %>%
  row_spec(6, bold = T) %>%
  row_spec(10, bold = T) %>%
  kable_styling()
```

\newpage

![**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species. Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b) show the distribution of resistance values for each drought.](tables_figures/Figure1.png){width=500px}


![**Figure 2. Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018. In these plots, heights are slightly offset for visualization purposes. Also shown are (d) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (e) $PLA_{dry}$ and (f) $\pi_{tlp}$. In (e-f), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals. In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}